---
title: "Baseline Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Baseline Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

The **Baseline method** estimates a patient's daily corticosteroid dose from
structured OMOP CDM fields — tablet strength, quantity dispensed, days supply,
and pre-computed `daily_dose` when available — using a four-step cascading
imputation strategy. No free-text SIG parsing is required.

This vignette demonstrates the full Baseline workflow on a synthetic dataset.

---

## Load package and synthetic data

```{r load}
library(SteroidDoseR)
library(dplyr)

extdata <- system.file("extdata", package = "SteroidDoseR")

drug_exp <- readr::read_csv(
  file.path(extdata, "synthetic_drug_exposure.csv"),
  show_col_types = FALSE
)

dplyr::glimpse(drug_exp)
```

---

## Step 1: Compute daily dose (Baseline method)

`calc_daily_dose_baseline()` tries four imputation methods in order and records
which one succeeded via the `imputation_method` column.

```{r baseline}
drug_baseline <- calc_daily_dose_baseline(drug_exp)

# Check imputation coverage
drug_baseline |>
  dplyr::count(imputation_method) |>
  dplyr::mutate(pct = round(100 * n / sum(n), 1))
```

---

## Step 2: Convert to prednisone equivalents

```{r conversion}
drug_baseline <- drug_baseline |>
  dplyr::mutate(drug_name_std = SteroidDoseR:::standardize_drug_name(drug_concept_name)) |>
  dplyr::rename(daily_dose_mg = daily_dose_mg_imputed)

drug_pe <- convert_pred_equiv(
  drug_baseline,
  drug_col = "drug_name_std",
  dose_col = "daily_dose_mg"
)

drug_pe |>
  dplyr::filter(pred_equiv_status == "ok") |>
  dplyr::select(person_id, drug_name_std, daily_dose_mg, pred_equiv_mg, equiv_factor) |>
  head(10)
```

---

## Step 3: Build episodes

```{r episodes}
# Keep only steroids with a computable dose
drug_for_episodes <- drug_pe |>
  dplyr::filter(pred_equiv_status == "ok", !is.na(pred_equiv_mg)) |>
  dplyr::rename(daily_dose_mg_imputed = pred_equiv_mg)

episodes <- build_episodes(
  drug_for_episodes,
  end_col  = "drug_exposure_end_date",
  gap_days = 30L
)

episodes
```

---

## Step 4: Summary statistics per episode

```{r summary}
episodes |>
  dplyr::group_by(person_id) |>
  dplyr::summarise(
    n_episodes        = dplyr::n(),
    total_days        = sum(n_days),
    mean_median_dose  = mean(median_daily_dose, na.rm = TRUE)
  )
```
