---
title: "OMOP Connector Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{OMOP Connector Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

SteroidDoseR Phase 2 introduces a **Connector** abstraction layer that lets
you point the package directly at an OMOP CDM database instead of manually
constructing a `drug_df`. The package handles SQL extraction, date filtering,
and field-availability detection automatically.

Two connector types are available:

| Connector | When to use |
|-----------|-------------|
| `create_omop_connector()` | Live OMOP CDM database (PostgreSQL, SQL Server, Snowflake, BigQuery, …) |
| `create_df_connector()` | In-memory data frame — tests, vignettes, synthetic data |

`convert_pred_equiv()` and `evaluate_against_gold()` remain data-frame–based
pure transformations and do not need a connector.

---

## Connect to OMOP CDM

### Real database (pseudocode — no credentials shown)

```r
library(SteroidDoseR)
library(DatabaseConnector)

# 1. Build connection details (DatabaseConnector standard)
cd <- createConnectionDetails(
  dbms     = "postgresql",        # or "sql server", "snowflake", "bigquery"
  server   = "myserver/omop_db",
  user     = "omop_reader",
  password = Sys.getenv("DB_PASSWORD"),
  port     = 5432
)

# 2. Wrap in a SteroidDoseR connector
con <- create_omop_connector(
  connectionDetails = cd,
  cdm_schema        = "cdm_531",   # schema with drug_exposure, concept, …
  vocab_schema      = "vocab",     # optional: separate vocabulary schema
  cdm_version       = "5.4"
)

# 3. (Optional) probe which drug_exposure fields are available at your site
con <- detect_capabilities(con)
con$capabilities
#> $has_sig        [1] TRUE
#> $has_days_supply [1] TRUE
#> $has_quantity    [1] TRUE
#> $has_route       [1] TRUE
```

The actual database connection is opened **only during the query** and closed
immediately after. There is no persistent connection to leak or time out.

---

## Demonstration with synthetic data (df_connector)

The examples below use `create_df_connector()` so they run without a database.
The code is identical to the real-database path — just swap
`create_df_connector(df)` for `create_omop_connector(cd, cdm_schema = "...")`.

```{r load-data}
library(SteroidDoseR)

extdata  <- system.file("extdata", package = "SteroidDoseR")
drug_exp <- readr::read_csv(
  file.path(extdata, "synthetic_drug_exposure.csv"),
  show_col_types = FALSE
)

# Wrap the data frame in a df_connector
con <- create_df_connector(drug_exp)
con
```

---

## One-call pipeline

`run_pipeline()` chains fetch → impute → convert → episode in a single call:

```{r run-pipeline}
episodes <- run_pipeline(
  con,
  method       = "baseline",  # or "nlp"
  return_level = "episode"    # or "exposure" for per-record output
)

episodes[, c("person_id", "drug_name_std", "episode_start",
             "episode_end", "n_days", "median_daily_dose")]
```

---

## Step-by-step connector workflow

For more control, call each function individually:

```{r step-by-step}
# Step 1: Baseline daily dose imputation
dose_df <- calc_daily_dose_baseline(con)

head(dose_df[, c("person_id", "drug_source_value",
                 "daily_dose_mg_imputed", "imputation_method")])

# Step 2: Prednisone-equivalent conversion (always data-frame based).
# Pass drug_concept_name directly — convert_pred_equiv() standardises names
# internally and falls back automatically when drug_name_std is absent.
equiv_df <- convert_pred_equiv(
  dose_df,
  drug_col = "drug_concept_name",
  dose_col = "daily_dose_mg_imputed"
)

head(equiv_df[, c("person_id", "drug_concept_name",
                  "daily_dose_mg_imputed", "pred_equiv_mg")])

# Step 3: Episode building
episodes2 <- build_episodes(
  equiv_df |> dplyr::filter(!is.na(pred_equiv_mg)),
  end_col  = "drug_exposure_end_date",
  dose_col = "pred_equiv_mg",
  gap_days = 30L
)
episodes2
```

---

## Filtering by person or concept

Both `run_pipeline()` and the individual functions accept filtering arguments
when a connector is supplied:

```r
# Pull only specific patients
doses <- calc_daily_dose_baseline(
  con,
  person_ids = c(1001L, 1002L, 1003L)
)

# Pull only specific drug concepts (OMOP concept IDs)
doses <- calc_daily_dose_nlp(
  con,
  drug_concept_ids = c(1518254L, 40224131L)  # prednisone, methylprednisolone
)
```

These arguments are silently ignored when `con` is a plain data frame, so
the same calling code works in both paths.

---

## Handling sites without a `sig` column

Some OMOP CDM instances do not populate `drug_exposure.sig`. Use
`sig_source = "drug_source_value"` to fall back to `drug_source_value` as the
SIG text source:

```r
doses <- calc_daily_dose_nlp(
  con,
  sig_source = "drug_source_value"
)
```

The package detects whether `sig` is absent or all-`NA` and aliases
`drug_source_value` automatically.

---

## Evaluate against a gold standard

After computing episodes via either method, pass the result to
`evaluate_against_gold()` as before:

```{r eval-gold}
gold_std <- readr::read_csv(
  file.path(extdata, "synthetic_gold_standard.csv"),
  show_col_types = FALSE
)

eval_result <- evaluate_against_gold(episodes, gold_std)
eval_result$summary[, c("coverage_pct", "MAE", "MBE", "RMSE")]
```

---

## Summary

| Step | Function | Connector-first? |
|------|----------|-----------------|
| Fetch + impute (structured) | `calc_daily_dose_baseline()` | Yes |
| Fetch + impute (NLP) | `calc_daily_dose_nlp()` | Yes |
| Convert to pred-equiv | `convert_pred_equiv()` | No (df only) |
| Build episodes | `build_episodes()` | Yes |
| Full pipeline | `run_pipeline()` | Yes |
| Evaluate | `evaluate_against_gold()` | No (df only) |