---
title: "Cross-Sectional (Point-in-Time) Comparison"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cross-Sectional (Point-in-Time) Comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Why this vignette exists

`evaluate_against_gold()` (see the `evaluation-workflow` vignette) is
**episode-level**: it matches a gold-standard episode -- a chart-reviewed
period that can span months -- against the best-overlapping *computed*
episode, which `build_episodes()` builds by gap-bridging prescriptions. If a
patient refills continuously with gaps under `gap_days` (default 30 days),
one computed episode can span the **entire study window**, years rather than
months. That is expected, not a bug -- episode-level evaluation was never
designed to answer "what dose was the patient on on a specific day."

That is a different, narrower question: **on a given office-visit encounter,
how do the methods compare for determining how much prednisone the patient
was on at that particular time?** This vignette walks through the tools that
answer it: `fetch_visit_occurrence()`, `dose_at_visits()`, and
`dose_agreement_metrics()`.

It is worth being explicit about what this does *not* do. Steroid
**trajectory** -- dose over time, or cumulative exposure -- is clinically
important and much harder to validate, because the gold standard only covers
discrete chart-reviewed intervals, not a continuous dose curve. Cross-sectional
comparison validates individual timepoints, which is a necessary building
block, but it does not itself validate a trajectory or cumulative-dose metric.
No gold standard currently exists to make that possible. See `bug.md`
(DECISION-13, DECISION-14) for the full reasoning.

---

## Prepare computed episodes (Baseline and NLP)

```{r setup-computed}
library(SteroidDoseR)
library(dplyr)

extdata  <- system.file("extdata", package = "SteroidDoseR")
drug_exp <- readr::read_csv(file.path(extdata, "synthetic_drug_exposure.csv"),
                            show_col_types = FALSE)
gold_std <- readr::read_csv(file.path(extdata, "synthetic_gold_standard.csv"),
                            show_col_types = FALSE)

baseline_episodes <- drug_exp |>
  calc_daily_dose_baseline() |>
  build_episodes(dose_col = "daily_dose_mg_imputed",
                 end_col  = "drug_exposure_end_date", gap_days = 30L)

nlp_episodes <- drug_exp |>
  calc_daily_dose_nlp() |>
  build_episodes(dose_col = "daily_dose_mg",
                 end_col  = "drug_exposure_end_date", gap_days = 30L)
```

Confirm episodes really do span long windows for at least one patient --
this is the behaviour that motivates a separate, shorter-window comparison:

```{r episode-spans}
baseline_episodes |>
  dplyr::mutate(n_days = as.integer(episode_end - episode_start) + 1L) |>
  dplyr::arrange(dplyr::desc(n_days)) |>
  dplyr::select(person_id, episode_start, episode_end, n_days) |>
  head(3)
```

---

## Fetch office-visit encounters

`fetch_visit_occurrence()` accepts either a live `omop_connector` or a plain
data frame already containing visit rows -- the synthetic fixture used here
takes the data-frame path, so no database is required:

```{r fetch-visits}
visits_raw <- readr::read_csv(file.path(extdata, "synthetic_visit_occurrence.csv"),
                              show_col_types = FALSE)

# visit_concept_id 9202 = OMOP standard "Outpatient Visit". This definition
# has no default in fetch_visit_occurrence() -- it must be confirmed per
# study, not silently guessed.
visits_df <- fetch_visit_occurrence(visits_raw, visit_concept_ids = c(9202L))
visits_df
```

---

## Look up the dose active at each visit

`dose_at_visits()` finds whichever episode(s) cover each visit date and sums
their dose (relevant when a patient is mid cross-titration between two
steroid drugs). Coverage semantics differ between computed methods and the
gold standard:

```{r dose-at-visits}
# Computed methods: no active episode == "not on steroids" (a real answer).
baseline_at_visits <- dose_at_visits(baseline_episodes, visits_df,
                                     visit_date_col = "visit_start_date")
nlp_at_visits <- dose_at_visits(nlp_episodes, visits_df,
                                visit_date_col = "visit_start_date")

# Gold standard: no coverage == "not chart-reviewed here," not "confirmed
# off steroids" -- use NA, not 0, and track has_coverage separately.
gold_at_visits <- dose_at_visits(gold_std, visits_df,
                                 visit_date_col    = "visit_start_date",
                                 dose_col          = "median_daily_dose",
                                 no_coverage_value = NA_real_)

cross_sectional_df <- tibble::tibble(
  person_id         = visits_df$person_id,
  visit_date        = visits_df$visit_start_date,
  baseline_dose_mg  = baseline_at_visits$dose_mg,
  nlp_dose_mg       = nlp_at_visits$dose_mg,
  gold_dose_mg      = gold_at_visits$dose_mg,
  has_gold_coverage = gold_at_visits$has_coverage
)
cross_sectional_df
```

---

## Agreement metrics

`dose_agreement_metrics()` computes the same MAE / MBE / RMSE / MAPE /
correlation formulas used inside `evaluate_against_gold()`, applied to plain
dose vectors -- so cross-sectional numbers stay comparable to episode-level
numbers.

```{r method-vs-method}
# Method vs. method, all visits (no_coverage_value = 0 makes "not on
# steroids" a comparable answer between methods).
dose_agreement_metrics(cross_sectional_df$baseline_dose_mg,
                       cross_sectional_df$nlp_dose_mg)
```

```{r method-vs-gold}
# Method vs. gold, restricted to visits that were actually chart-reviewed.
covered <- cross_sectional_df |> dplyr::filter(has_gold_coverage)

dose_agreement_metrics(covered$baseline_dose_mg, covered$gold_dose_mg)
```

---

## Approximating trajectory truth: carry-forward assumption

Gold coverage is sparse -- only `r sum(gold_at_visits$has_coverage)` of
`r nrow(visits_df)` visits above were actually chart-reviewed. The PI
proposed a pragmatic mitigation for the still-open trajectory gold-standard
gap (see `bug.md` DECISION-14): assume a patient's dose stays constant from
one office visit to the next, and carry the last *actually reviewed* gold
dose forward (last-observation-carried-forward, never backward) to fill
unreviewed visits.

**This is an assumption, not a fact.** A patient's dose can genuinely change
between visits without being caught by chart review -- most acutely during
tapers, which is exactly when the dose is *expected* to change. For that
reason `carry_forward_dose()` always tags provenance with a `dose_source`
column, and accuracy must always be reported twice: once for `"reviewed"`
visits only (identical to the numbers above), once including
`"carried_forward"` visits (larger n, but resting on the assumption). Never
blend the two into a single number.

```{r carry-forward}
gold_at_visits_cf <- gold_at_visits |>
  carry_forward_dose(visit_date_col = "visit_start_date",
                     dose_col       = "dose_mg",
                     coverage_col   = "has_coverage")

table(gold_at_visits_cf$dose_source)
```

```{r carry-forward-accuracy}
# reviewed_only -- byte-identical to the method-vs-gold result above
reviewed <- gold_at_visits_cf$dose_source == "reviewed"
dose_agreement_metrics(baseline_at_visits$dose_mg[reviewed],
                       gold_at_visits_cf$dose_mg[reviewed])

# reviewed_plus_carried_forward -- larger n, rests on the assumption
covered_cf <- gold_at_visits_cf$dose_source %in% c("reviewed", "carried_forward")
dose_agreement_metrics(baseline_at_visits$dose_mg[covered_cf],
                       gold_at_visits_cf$dose_mg[covered_cf])
```

Note the second call typically shows *worse* accuracy than the first -- that
is the expected, honest effect of exposing the assumption to more data
points, not a bug.

---

## Running this automatically

`CodeToRun.R` (STEP 2c) and `CompareToRun.R` (section 11b) run this same
sequence automatically once `VISIT_CONCEPT_IDS` is set in `RunAll.R`,
including the carry-forward step, writing `cross_sectional_comparison.csv`
(with a `dose_source` column) and `cross_sectional_summary.csv` (with a
`truth_scope` column) alongside the existing episode-level outputs.
