---
title: "Evaluation vs Gold Standard"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Evaluation vs Gold Standard}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

`evaluate_against_gold()` matches computed episodes to a manually-reviewed
gold standard using a date-overlap join, then returns coverage (% usable
episodes), MAE, and MBE — the three primary metrics from the OHDSI 2025
AgentDose poster — plus a richer stratified breakdown.

This vignette uses the synthetic gold standard that ships with the package.

---

## Prepare computed episodes (using NLP method)

```{r setup-computed}
library(SteroidDoseR)
library(dplyr)

extdata  <- system.file("extdata", package = "SteroidDoseR")
drug_exp <- readr::read_csv(file.path(extdata, "synthetic_drug_exposure.csv"),
                            show_col_types = FALSE)
gold_std <- readr::read_csv(file.path(extdata, "synthetic_gold_standard.csv"),
                            show_col_types = FALSE)

# Run NLP pipeline
drug_nlp <- calc_daily_dose_nlp(drug_exp)
drug_pe  <- convert_pred_equiv(drug_nlp, drug_col = "drug_name_std",
                                          dose_col = "daily_dose_mg")

computed_episodes <- drug_pe |>
  dplyr::filter(!is.na(pred_equiv_mg)) |>
  dplyr::rename(daily_dose_mg_imputed = pred_equiv_mg) |>
  build_episodes(end_col = "drug_exposure_end_date", gap_days = 30L) |>
  dplyr::rename(median_daily_dose = median_daily_dose)
```

---

## Run evaluation

```{r evaluate}
results <- evaluate_against_gold(computed_episodes, gold_std)
```

---

## Primary metrics (poster framing)

```{r metrics}
s <- results$summary
cat(sprintf("Usable / computable episodes: %d / %d (%.1f%%)\n",
            s$n_matched_periods, s$n_gold_periods, s$coverage_pct))
cat(sprintf("MAE  = %.2f mg/day\n", s$MAE))
cat(sprintf("MBE  = %.2f mg/day  (positive = over-estimation)\n", s$MBE))
cat(sprintf("RMSE = %.2f mg/day\n", s$RMSE))
cat(sprintf("Pearson r = %.3f  |  Spearman rho = %.3f\n",
            s$pearson_corr, s$spearman_corr))
```

---

## Agreement categories

```{r agreement}
results$comparison |>
  dplyr::count(agreement_category) |>
  dplyr::mutate(pct = round(100 * n / sum(n), 1)) |>
  dplyr::arrange(match(agreement_category,
    c("Exact (<=5%)", "Good (<=20%)", "Moderate (<=50%)", "Poor (>50%)")))
```

---

## Stratified analysis by dose range

```{r strat-dose}
results$stratified$by_dose_range
```

---

## Per-episode detail

```{r detail}
results$comparison |>
  dplyr::select(patient_id, episode_start, episode_end,
                gold_dose, computed_dose, absolute_error, agreement_category) |>
  head(10)
```
