---
title: "Clinical Note NLP Workflow (medspaCy)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Clinical Note NLP Workflow (medspaCy)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE   # Python required; skip during R CMD CHECK
)
```

## Overview

The medspaCy NLP method extracts steroid doses from **unstructured clinical note
text** — physician assessments, discharge summaries, progress notes — where
structured SIG strings are absent or incomplete.

The pipeline runs in five steps:

1. **Regex SIG parser** (`parse_sig_one_advanced`) on the `sig` column — fast path.
2. **medspaCy** — fires only for records the regex could not resolve.
3. **Strength fallback** — `amount_value` → drug concept name.
4. **Baseline structural fallback** (M1/M3/M4).
5. **Plausibility cap** (2 000 mg/day default).

## Python setup

Install once in your Python environment:

```bash
pip install medspacy scispacy
python -m spacy download en_core_sci_sm
```

Then point `reticulate` at that environment before loading `SteroidDoseR`:

```r
library(reticulate)
# use_virtualenv("~/.venv/steroids")   # if using a virtualenv
# use_condaenv("steroids")             # if using conda
```

## Basic usage

```{r basic}
library(SteroidDoseR)

df <- tibble::tibble(
  person_id                = 1L,
  drug_concept_name        = "prednisone 5 MG oral tablet",
  route_concept_name       = "oral",
  sig                      = NA_character_,
  clinical_note            = paste(
    "Assessment: Patient continued on prednisone 40 mg daily.",
    "Plan: Taper to 20 mg over the next 4 weeks."
  ),
  drug_exposure_start_date = as.Date("2023-01-01"),
  drug_exposure_end_date   = as.Date("2023-06-01")
)

result <- calc_daily_dose_nlp_notes(df, note_col = "clinical_note")

result[, c("person_id", "daily_dose_mg", "parsed_status",
           "note_section", "note_negation_flag", "note_uncertainty_flag")]
```

## Parsing a single note

```{r single}
parse_note_one("Patient on methylprednisolone 1 g IV daily for 3 days.")
parse_note_one("Patient is NOT currently on any corticosteroids.")
parse_note_one("Consider starting prednisone 40 mg if flare worsens.")
```

| Input | `parsed_status` | `note_negation_flag` | `note_uncertainty_flag` |
|---|---|---|---|
| "...on prednisone 40 mg daily" | `notes_ok` | FALSE | FALSE |
| "NOT on any corticosteroids" | `notes_negated` | TRUE | FALSE |
| "Consider starting prednisone" | `notes_uncertain` | FALSE | TRUE |
| "was previously on prednisone" | `notes_historical` | FALSE | FALSE |

## Note-specific output columns

| Column | Type | Meaning |
|---|---|---|
| `note_entity_count` | int | Number of steroid entities found |
| `note_negation_flag` | lgl | Drug is negated in the note |
| `note_uncertainty_flag` | lgl | Drug is hedged / uncertain |
| `note_section` | chr | Detected section (`"assessment"`, `"plan"`, `"history"`, …) |

## Hybrid: SIG + note in the same data frame

When the `sig` column is populated for some records and the `clinical_note`
column for others, `calc_daily_dose_nlp_notes()` handles both automatically:

```{r hybrid}
df_mixed <- tibble::tibble(
  person_id                = 1:3,
  drug_concept_name        = rep("prednisone 5 MG oral tablet", 3),
  route_concept_name       = rep("oral", 3),
  sig = c(
    "Take 2 tablets (10 mg) daily",   # resolved by regex
    NA_character_,                     # falls through to medspaCy
    NA_character_                      # falls through to medspaCy
  ),
  clinical_note = c(
    NA_character_,
    "Patient on prednisone 20 mg daily.",
    "No steroids prescribed at this visit."
  ),
  drug_exposure_start_date = rep(as.Date("2023-01-01"), 3),
  drug_exposure_end_date   = rep(as.Date("2023-06-01"), 3)
)

result_mixed <- calc_daily_dose_nlp_notes(df_mixed, note_col = "clinical_note")
result_mixed[, c("person_id", "daily_dose_mg", "parsed_status")]
```

## Full pipeline

```{r pipeline}
episodes <- run_pipeline(
  df_mixed,
  method    = "nlp_notes",
  note_col  = "clinical_note",
  gap_days  = 30L
)
episodes[, c("person_id", "drug_name_std", "episode_start",
             "episode_end", "median_daily_dose")]
```

## Inspecting negated and uncertain records

Negated or uncertain records will have `NA` for `daily_dose_mg` from the note
step and will fall through to the baseline structural fallback. You can inspect
them directly:

```{r flags}
library(dplyr)

result |>
  filter(note_negation_flag | note_uncertainty_flag) |>
  select(person_id, clinical_note, parsed_status,
         note_negation_flag, note_uncertainty_flag, daily_dose_mg)
```

## Converting to prednisone equivalents and building episodes

The output of `calc_daily_dose_nlp_notes()` feeds directly into the rest of
the standard pipeline:

```{r convert}
library(SteroidDoseR)

notes_result <- calc_daily_dose_nlp_notes(df, note_col = "clinical_note")

# Convert to prednisone equivalents
pe_result <- convert_pred_equiv(notes_result, dose_col = "daily_dose_mg")

# Build episodes
episodes <- build_episodes(pe_result, dose_col = "pred_equiv_mg", gap_days = 30L)
```
