---
title: "NLP Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NLP Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

The **NLP method** uses rule-based parsing of free-text SIG (prescription
instruction) strings to extract dose, frequency, and duration information.
It handles common abbreviations (BID, TID, QID, QOD), explicit mg totals,
and taper language. Records with unstructured SIGs are flagged via
`parsed_status`.

This vignette demonstrates the full NLP workflow on the same synthetic dataset
used in the Baseline vignette.

---

## Load data

```{r load}
library(SteroidDoseR)
library(dplyr)

extdata  <- system.file("extdata", package = "SteroidDoseR")
drug_exp <- readr::read_csv(
  file.path(extdata, "synthetic_drug_exposure.csv"),
  show_col_types = FALSE
)
```

---

## Step 1: Parse SIG strings (NLP)

`calc_daily_dose_nlp()` filters to oral systemic steroids, standardises drug
names, and parses each SIG. Records for non-steroid drugs (e.g. hydroxychloroquine)
are dropped automatically.

```{r nlp}
drug_nlp <- calc_daily_dose_nlp(
  drug_exp,
  drug_name_col = "drug_concept_name",
  sig_col       = "sig"
)

# Coverage report
drug_nlp |>
  dplyr::count(parsed_status) |>
  dplyr::mutate(pct = round(100 * n / sum(n), 1))
```

---

## Step 2: Inspect parsed components

```{r inspect}
drug_nlp |>
  dplyr::filter(parsed_status == "ok") |>
  dplyr::select(person_id, drug_name_std, sig, tablets,
                freq_per_day, mg_per_admin, daily_dose_mg) |>
  head(10)
```

---

## Step 3: Convert to prednisone equivalents

```{r convert}
drug_nlp_pe <- convert_pred_equiv(
  drug_nlp,
  drug_col = "drug_name_std",
  dose_col = "daily_dose_mg"
)

drug_nlp_pe |>
  dplyr::filter(pred_equiv_status == "ok") |>
  dplyr::select(person_id, drug_name_std, daily_dose_mg,
                pred_equiv_mg, equiv_factor) |>
  head(8)
```

---

## Step 4: Build episodes

```{r episodes}
drug_for_ep <- drug_nlp_pe |>
  dplyr::filter(pred_equiv_status == "ok", !is.na(pred_equiv_mg)) |>
  dplyr::rename(daily_dose_mg_imputed = pred_equiv_mg)

episodes <- build_episodes(
  drug_for_ep,
  end_col  = "drug_exposure_end_date",
  gap_days = 30L
)

episodes
```

---

## Step 5: Parseable coverage summary

```{r coverage}
total   <- nrow(drug_nlp)
ok_rows <- sum(drug_nlp$parsed_status == "ok", na.rm = TRUE)

cat(sprintf(
  "Parseable records: %d / %d (%.1f%%)\n",
  ok_rows, total, 100 * ok_rows / total
))
```
